---
title: "Introduction to R"
author: R.Slieker
output: github_document
email: r.slieker at vumc.nl
---


## Where to get this page from: 
http://goo.gl/cWAAyv


## Introduction


## The R interface
- Originated from proprietary S programming language

- Comes in many flavours, like GUI (hardly a user interface), Rstudio, Eclipse

- Save scripts as R files, can be opened with all kinds of software (R, Rstudio, sublime, Notepad etc. )

- Save objects to Rdata files, so you can continue with it next time without the need for rerunning the complete script

## Other software to be used with R

- Rstudio is very useful: https://www.rstudio.com/products/RStudio/
It has many advantages: integrated help, use of tab, 

- Sublime when only typing syntax (to link with servers): https://www.sublimetext.com/

- Packages: for an (almost infinity) number of applications people have developed packages, for example Bioconductor: http://bioconductor.org/

- Git to track your code: https://git.lumc.nl 


## Suggested literature:
  
*The art of R programming, Normal Matlof, Starch Press*

*Introductory Statistics with R, Peter Dalgaard, Springer*



## Starting simple with R

```{r start}
1+1

#Assign values to variables
a <- 1
a + a

#But the equal sign also works!
aa = 1
aa+aa
#But professionals use the arrow and not the equal sign to distinguish from  == :)

#Or multiple values ot variables
b <- c(1,2,3) # c, concatenate
b
```



## The main classes
- Numeric and integers
- Characters
- Factors

```{r classes}
#Numeric
num <- c(1,2,3)
#Class can be used to view the classs
class(num)

#Character
characters <- c("A","B","C")
class(characters)

#Factor, same as character but with the levels in the data
factors <- factor(c("A","A","B","B","C"), levels=c("A","B","C"))
class(factors)
```

## Objects
- Data frames
  
  Advantages: can hold multiple classes, columns can be viewed easily
  Disadvantage: one cannot do analyses on the entire object
  

- Matrices
  
  Advantage: very quick and useful for matrix calculations.
  Disadvantage: less easy to modify (add columns, change class of data)

```{r objects}
#Data.frame: can hold multiple classes
df <- data.frame(Group = c(1,2,3), Outcome = c(2,3,6))

#head can be used to view objects
head(df)
#str will give the classes of the different columns
str(df)
class(df)

#Groups can called with the dollar sign (dataframe only)
df$Group


#Matrix:
mtrx <- matrix(1:25, nrow=5)
head(mtrx)
str(mtrx)
class(mtrx)

#Calculate something at once
mtrx*mtrx
```

## Modifying objects

```{r modifying}
#Adding rownames to files 
rownames(df) <- c("A","B","C")
#The file has already colnames
colnames(df)

#What doesn't work:
dim(mtrx)
#rownames(mtrx) <- c("A","B","C")
#Why not?

## Learn to understand error codes (read them carefully)

## If you don't get the error, google it. There is ALWAYS someone before you who had the same question.


## This does work. Why?
rownames(mtrx) <- c("A","B","C","D","E")

#Data.frame
df[1:2, ] #First two lines

#Add column
df$NewCol <- c(1,2,5)
head(df)

#Matrix:
mtrx[1:2,]
mtrx[1:2] #Different!

# List 
list(GroupA = 1:3, GroupB = 2:5 , GroupC = 3:5)
```

## Reading from files

R can read all types of files (with or without required packages), such as txt, csv, xlsx (package: xlsx), sav spss file  (package: Hmisc)

```{r read}
# Setting a working directory (= tell R where to find files and where to save files.)
setwd("/Users/roderick/Documents/Onderwijs/IntroR/")

#Reading csv files (comma delimited)
IrisCSV <- read.table("Iris.csv", sep=",", header=T)

#Or we give the full path
IrisCSV <- read.table("/Users/roderick/Documents/Onderwijs/IntroR/Iris.csv", sep=",", header=T)

head(IrisCSV)
str(IrisCSV)

#Reading tab delimited 
IrisTXT <- read.table("Iris.txt", sep="\t", header=T)
head(IrisTXT)
str(IrisTXT)
```

## Summary statistics 

There are various ways of calcalating summary statistics. A useful function is **by**.

```{r summaries}

# Means
# by(data = someVariable, INDICES = someGrouping, FUN = aGroupingFunction)

by(data = IrisCSV$Sepal.Length, INDICES = IrisCSV$Species, mean)
# Same as:
by(IrisCSV$Sepal.Length, IrisCSV$Species, mean)
#Store in variable
meanGroup <- by(data = IrisCSV$Sepal.Length, INDICES = IrisCSV$Species, mean)
class(meanGroup)
#Small trick:
as.list(meanGroup)


```


## Packages

Packages are an infinite rich source to facilitate your analysis

*To install packages from CRAN*

install.packages("ggplot2")

*Or when from Bioconductor:*

source("https://bioconductor.org/biocLite.R")

biocLite("IRanges")


## Plotting
R base (without packages) can plot all kind of plots


```{r plotting}
library(ggplot2)

#Scatter plot
plot(IrisCSV$Sepal.Length, IrisCSV$Sepal.Width, xlab="Sepal length", ylab="Sepal width")

ggplot(IrisCSV, aes(x=Sepal.Length, y=Sepal.Width, col=Species))+
  geom_point()+
  ggtitle("Some title")+
  xlab("Sepal length")+
  ylab("Sepal width")+
  scale_color_manual(values = c("#009AC7","#132B41","#8B1A4F"))
  


# Boxplot
boxplot(IrisCSV$Sepal.Length~IrisCSV$Species, xlab="Sepal length", ylab="Sepal width")

ggplot(IrisCSV, aes(x=Species, y=Sepal.Length, fill=Species))+ #Define variables
  geom_boxplot()+ #What kind of graph?
  ggtitle("Some title")+ #Add title to graph
  xlab("Sepal length")+ #Add x-axis labels
  ylab("Sepal width")+ #Add y-axis labels
  scale_fill_manual(values = c("#009AC7","#132B41","#8B1A4F")) # Change colors using HEX colors

# Smoothed line graph
ggplot(IrisCSV, aes(x=Sepal.Width, y=Sepal.Length, col=Species))+
  geom_point()+
  geom_smooth(method=lm)+
  ggtitle("Some title")+
  xlab("Sepal length")+
  ylab("Sepal width")+
  scale_color_manual(values = c("#009AC7","#132B41","#8B1A4F"))

ggplot(IrisCSV, aes(x=Sepal.Width, y=Sepal.Length, col=Species))+
  geom_point()+
  geom_smooth(method=loess)+
  ggtitle("Some title")+
  xlab("Sepal length")+
  ylab("Sepal width")+
  scale_color_manual(values = c("#009AC7","#132B41","#8B1A4F"))

```






