---
title: "Introduction to R"
output:
  github_document:
    toc: true
    toc_depth: 2 
---

This page can be found at:  
http://goo.gl/cWAAyv

#1.Introduction#

## The R interface##
- Originated from proprietary S programming language

- Comes in many flavours, like GUI (hardly a user interface), Rstudio, Eclipse

- Save scripts as R files, can be opened with all kinds of software (R, Rstudio, sublime, Notepad etc. )

- Safer and more reproducible than other software: source data is not changed, scripts can be shared with others

- Save objects to Rdata files, so you can continue with it next time without the need for rerunning the complete script


##Other software to be used with R ##

- Rstudio is very useful: https://www.rstudio.com/products/RStudio/

It has many advantages: integrated help, use of tab key, more stable than Rgui. 

- Sublime when only typing syntax (to link with servers): https://www.sublimetext.com/

- Packages: for an (almost infinity) number of applications people have developed packages, for example Bioconductor: http://bioconductor.org/

- Git to track your code: https://git.lumc.nl 


## Suggested literature:##
  
*Google is your best friend. Every question has been asked more than once by someone else*

*The art of R programming, Normal Matlof, Starch Press*

*Introductory Statistics with R, Peter Dalgaard, Springer*  



# 2. Starting simple with R

```{r start}
1+1

#Assign values to variables
a <- 1
a + a

#But the equal sign also works!
aa = 1
aa+aa
#But professionals use the arrow and not the equal sign to distinguish from  == :)

#Or multiple values ot variables
b <- c(1,2,3) # c, concatenate
b

#Variables can have any names, but don't assign names to existing objects, such as sum, c, sd etc
#Spaces are not allowed as well as special symbols


#Sequences
rep(x = 1,times = 5)
seq(from = 1,to = 5,by=0.5)
seq(from = 1,to = 5,length.out = 100)
```



#3. The main classes
- Numeric and integers
- Characters
- Factors
- Lists

```{r classes}
#Numeric
num <- c(1,2,3)
#Class can be used to view the classs
class(num)

#Character
characters <- c("A","B","C")
class(characters)

#Factor, same as character but with the levels in the data
factors <- factor(c("A","A","B","B","C"), levels=c("A","B","C"))
class(factors)


# List 
list(GroupA = 1:3, GroupB = 2:5 , GroupC = 3:5)
```

#4. Objects 

##Creating objects##

- Data frames
  
  Advantages: can hold multiple classes, columns can be viewed easily
  Disadvantage: one cannot do analyses on the entire object
  

- Matrices
  
  Advantage: very quick and useful for matrix calculations.
  Disadvantage: less easy to modify (add columns, change class of data)

```{r objects}
#Data.frame: can hold multiple classes
df <- data.frame(Group = c(1,2,3), Outcome = c(2,3,6))

#head can be used to view objects
head(df)
#str will give the classes of the different columns
str(df)
class(df)

#Groups can called with the dollar sign (dataframe only)
df$Group


#Matrix:
mtrx <- matrix(1:25, nrow=5)
head(mtrx)
str(mtrx)
class(mtrx)

#Calculate something at once
mtrx*mtrx
```

##Modifying objects##

```{r modifying}
#Adding rownames to files 
rownames(df) <- c("A","B","C")
#The file has already colnames
colnames(df)

#What doesn't work:
#rownames(mtrx) <- c("A","B","C")
#Why not?
## Learn to understand error codes (read them carefully)
## If you don't get the error, google it. There is ALWAYS someone before you who had the same question.
## This does work. Why?
rownames(mtrx) <- c("A","B","C","D","E")




#Modifying objects

#Data.frame
df[1:2, ] #First two lines

#Different (order has changed!)
df[2:1, ] 

#You can also subset by rownames
df[c("A","B"),]

#Or colnames
df[,c("Group","Outcome")]

#Add new columns
df$NewCol <- c(1,2,5)
head(df)


# Modifying a matrix:
mtrx[1:2,]
mtrx[1:2] #Different!
```

#5. Loading and saving from R

##Reading from files##

R can read all types of files (with or without required packages), such as txt, csv, xlsx (package: xlsx), sav spss file  (package: Hmisc)

```{r read}
# Setting a working directory (= tell R where to find files and where to save files.)
setwd("/Users/roderick/Documents/Onderwijs/IntroR/")

#Reading csv files (comma delimited)
IrisCSV <- read.table("Iris.csv", sep=",", header=T)

#Or we give the full path
IrisCSV <- read.table("/Users/roderick/Documents/Onderwijs/IntroR/Iris.csv", sep=",", header=T)

head(IrisCSV)
str(IrisCSV)

#Reading tab delimited 
IrisTXT <- read.table("Iris.txt", sep="\t", header=T)
head(IrisTXT)
str(IrisTXT)
```

##Save to files##

```{r save}
data(iris)

# Save objects in analysis
save(iris, file="Iris.RData")
# Load objects from previous sessions
load("Iris.RData")

#One can also save to common formats such as csv, txt
write.table(iris, file="Iris as text file.txt", sep="\t", col.names=T, row.names=F)
write.table(iris, file="Iris as csv file.csv", sep=";", col.names=T, row.names=F)
```

#6. Describing data using descriptive statistics

```{r summaries}
#The iris dataset is a build in dataset

#Describe dataset
nrow(iris)
ncol(iris)
length(iris$Sepal.Length)
sum(iris$Sepal.Length)
mean(iris$Sepal.Length)
median(iris$Sepal.Length)
sd(iris$Sepal.Length)

# Summary statistics by group
# by(data = someVariable, INDICES = someGrouping, FUN = aGroupingFunction)

by(data = iris$Sepal.Length, INDICES = iris$Species, mean)
# Same as:
by(iris$Sepal.Length, iris$Species, mean)
#Store in variable
meanGroup <- by(data = iris$Sepal.Length, INDICES = iris$Species, FUN = mean)
SDGroup <- by(data = iris$Sepal.Length, INDICES = iris$Species, FUN = sd)
LengthGroup <- by(data = iris$Sepal.Length, INDICES = iris$Species, FUN = length)

class(meanGroup)

#Small trick:
as.list(meanGroup)
statistics <- cbind( as.list(meanGroup),as.list(SDGroup), as.list(LengthGroup))
colnames(statistics) <- c("Mean","SD","n")
statistics
```


#7.Packages

Packages are an infinite rich source to facilitate your analysis

*To install packages from CRAN*

install.packages("ggplot2")

*Or when from Bioconductor:*

source("https://bioconductor.org/biocLite.R")

biocLite("IRanges")


#8.Plotting
R base (without packages) can plot all kind of plots but are not very nice. 

ggplot2 is a very commonly used package for plotting. Although the approach is a bit different, the plots are nicer and more flexible to change. For more help see: http://ggplot2.tidyverse.org/reference/ or google!


```{r plotting}
library(ggplot2)

#Scatter plot
plot(iris$Sepal.Length, iris$Sepal.Width, xlab="Sepal length", ylab="Sepal width")

ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, col=Species))+ # Define variables
  geom_point()+ # We want to plot dots
  ggtitle("Some title")+ #Add title to graph
  xlab("Sepal length")+ #Add x-axis labels
  ylab("Sepal width")+ #Add y-axis labels
  scale_color_manual(values = c("#009AC7","#132B41","#8B1A4F"))


# Boxplot
boxplot(iris$Sepal.Length~IrisCSV$Species, xlab="Sepal length", ylab="Sepal width")

ggplot(iris, aes(x=Species, y=Sepal.Length, fill=Species))+ #Define variables
  geom_boxplot()+ #What kind of graph?
  ggtitle("Some title")+ #Add title to graph
  xlab("Sepal length")+ #Add x-axis labels
  ylab("Sepal width")+ #Add y-axis labels
  scale_fill_manual(values = c("#009AC7","#132B41","#8B1A4F")) # Change colors using HEX colors

# Jitter plot
ggplot(iris, aes(x=Species, y=Sepal.Length, col=Species))+ #Define variables
  geom_jitter(width=0.2)+
  ggtitle("Some title")+ 
  xlab("Sepal length")+ 
  ylab("Sepal width")+ 
  scale_colour_manual(values = c("#009AC7","#132B41","#8B1A4F")) 

# Smoothed line graph
ggplot(IrisCSV, aes(x=Sepal.Width, y=Sepal.Length, col=Species))+
  geom_point()+
  geom_smooth(method=lm)+
  ggtitle("Some title")+
  xlab("Sepal length")+
  ylab("Sepal width")+
  scale_color_manual(values = c("#009AC7","#132B41","#8B1A4F"))

ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, col=Species))+
  geom_point()+
  geom_smooth(method=loess)+
  ggtitle("Some title")+
  xlab("Sepal length")+
  ylab("Sepal width")+
  scale_color_manual(values = c("#009AC7","#132B41","#8B1A4F"))

```

##Saving plots##
```{r savingplots}
library(ggplot2)

pdf("Boxplot of iris data.pdf")
ggplot(iris, aes(x=Species, y=Sepal.Length, fill=Species))+ #Define variables
  geom_boxplot()+ #What kind of graph?
  ggtitle("Some title")+ #Add title to graph
  xlab("Sepal length")+ #Add x-axis labels
  ylab("Sepal width")+ #Add y-axis labels
  scale_fill_manual(values = c("#009AC7","#132B41","#8B1A4F")) # Change colors using HEX colors
dev.off()

jpeg("Boxplot of iris data.jpg", width = 10, height = 10, units="cm", res=200)
ggplot(iris, aes(x=Species, y=Sepal.Length, fill=Species))+ #Define variables
  geom_boxplot()+ #What kind of graph?
  ggtitle("Some title")+ #Add title to graph
  xlab("Sepal length")+ #Add x-axis labels
  ylab("Sepal width")+ #Add y-axis labels
  scale_fill_manual(values = c("#009AC7","#132B41","#8B1A4F")) # Change colors using HEX colors
dev.off()

```



