---
title: "Introduction to R"
output: github_document
---


## Where to get this page from: 


## Beginning simple

- Originated from proprietary S programming language

- Comes in many flavours, like GUI (hardly a user interface), Rstudio, Eclipse

- Rstudio is very useful: https://www.rstudio.com/products/RStudio/

- Sublime when only typing syntax (to link with servers): https://www.sublimetext.com/

- Git to track your code: https://git.lumc.nl 

- Packages: for an (almost infinity) number of applications people have developed packages, for example Bioconductor: http://bioconductor.org/

- Suggested literature:


```{r start}
1+1

#Assign values to variables
a <- 1
a + a

#Or multiple values ot variables
b <- c(1,2,3) # c, concatenate
b
```



## String classes
- Numeric
- Characters
- Factors

```{r classes}
#Numeric
num <- c(1,2,3)
#Class can be used to view the classs
class(num)

#Character
characters <- c("A","B","C")
class(characters)

#Factor
factors <- factor(c("A","A","B","B","C"), levels=c("A","B","C"))
class(factors)
```

## Objects
- Data frames
  
  Advantages: can hold multiple classes, columns can be viewed easily
  Disadvantage: one cannot do analyses on the entire object
  

- Matrices
  
  Advantage: very quick and useful for matrix calculations.
  Disadvantage: less easy to modify (add columns)

```{r objects}
#Data.frame: can hold multiple classes
df <- data.frame(Group = c(1,2,3), Outcome = c(2,3,6))

#head can be used to view objects
head(df)
#str will give the classes of the different columns
str(df)
class(df)

#Groups can called with the dollar sign (dataframe only)
df$Group


#Matrix:
mtrx <- matrix(1:25, nrow=5)
head(mtrx)
str(mtrx)
class(mtrx)
```

## Modifying objects

```{r modifying}
#Data.frame
df[1:2, ] #First two lines

#Add column
df$NewCol <- c(1,2,5)
head(df)

#Matrix:
mtrx[1:2,]
mtrx[1:2] #Different!

```

## Reading from files

R can read all types of files (with or without required packages), such as txt, csv, xlsx (package: xlsx), sav spss file  (package: Hmisc)

```{r read}
#read.table(" ")

```




#Plotting: mult
```{r plotting}

```






